---
title: "Graphs for Communication - Creating Animated Maps "
author:
    name: Chris Lee
    url: https://clfee.github.io/
date: 2024-09-01
format: 
  html:
        html-math-method: katex
        grid: 
            margin-width: 350px
fig-width: 8
fig-height: 6
editor_options: 
  chunk_output_type: console
execute: 
  echo: fenced
reference-location: margin
citation-location: margin
categories:
  - R 
  - Data organization
---

Making the data easier to understand!

**Goal :** To demonstrate how to create animated maps with `ggplot2` and `gganimate` in R to track MRSA BSI incidence in the State of California.

## Graphs for Communication

Earlier this year, I was preparing a presentation for stakeholders to showcase how our business performed, especially after large events and gatherings were no longer restricted. While we saw solid growth overall, it wasn’t consistent across all locations. One key focus of the presentation was to highlight how sales evolved over time at wedding venues and restaurants.

I wanted to present the data in a way that was not only accurate but also engaging enough to capture everyone's attention. To achieve this, I turned to the `ggplot2` and `gganimate` packages in R. These tools allowed me to transform dry numbers into visually compelling animations, making the data easier to understand and far more impact for my audience.

The same storytelling strategy can also be applied to the datasets of Methicillin-resistant Staphylococcus aureus (MRSA) bloodstream infections (BSI) in [California Hospitals](https://data.chhs.ca.gov/dataset/methicillin-resistant-staphylococcus-aureus-mrsa-bloodstream-infections-bsi-in-california-hospitals).

The datasets include the associated 95% confidence intervals for the SIR and statistical interpretation to show whether MRSA BSI incidence was the same (no different), better (lower) or worse (higher) than the national baseline. Another performance measure in this dataset allows for tracking hospital progress in meeting national HAI reduction goals. Hospitals must have an SIR at or below incremental targets each year to be considered on track.

In this practice, we will generate an animated map displays whether MRSA BSI incidence in the State of California was the same (no different), better (lower) or worse (higher) than the national baseline from 2019 – 2023. The comparison is depicted values and colors range from -1 (purple, worse than the national average) to 1 (yellow, better than the national average).

## Steps

The first section simply loads the libraries and data that will be used.

```{r}
#| context: setup
#| code-fold: true
#| warning: false

# Load required packages
library(gganimate)
library(gifski)
library(dplyr)
library(tidyverse)
library(dplyr)
library(ggplot2)


# Load data
url_2022 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/3ff809c6-c1e2-4107-be0c-f976521aadb2/download/mrsa_bsi_odp_2022.csv"
url_2021 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/23bef156-ae4d-4800-8512-d29660b2269f/download/cdph_mrsa_bsi_odp_2021.csv"
url_2019 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/62aead61-e320-4a74-b64c-1f65efa72b35/download/cdph_mrsa_bsi_odp_2019.csv"
url_2023 <-"https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/15c4962f-e357-4b03-9d24-655a2bc8030c/download/mrsa_bsi_odp_2023.csv"

data_2023 <- read_csv(url_2023)
data_2022 <- read_csv(url_2022)
data_2021 <- read_csv(url_2021)
data_2019 <- read_csv(url_2019)

# Combine data
mrsa_combine <- bind_rows(data_2023,data_2022, data_2021, data_2019)

# Remove redundant data
mrsa_combine <- unique(mrsa_combine)
# Lon and lat info
us_counties <- map_data("county")|> filter(region =="california")|> mutate( subregion = toupper( subregion))|> dplyr::select(-c( order,  region)) 

# Data transformation
mrsa_combine <- mrsa_combine |>
  dplyr::select(-State, -HAI, -Met_2020_Goal, -SIR_CI_95_Lower_Limit,-SIR_CI_95_Upper_Limit, -On_Track, -Notes) |>
  mutate(Months = as.numeric(ifelse(is.na(Months), 13, Months)),
         Year = as.integer(ifelse(is.na(Year), 2021, Year)),
         Comparison = case_when(
      Comparison == "Worse" ~ -1,
      Comparison == "Better" ~ 1,
      Comparison == "Same" ~ 0,
      .default = 0),
         Patient_Stay = cut(      
           Patient_Days,
           breaks = quantile(Patient_Days, probs = seq(0, 1, by = 0.1), na.rm = TRUE),
           labels = 1:10),
         Months = as.character(Months),
         Year = as.character(Year),
         Infections_Reported = as.numeric(Infections_Reported))|>
  filter(!grepl("Pooled", Facility_Name, ignore.case = TRUE),
         !grepl("Pooled", Facility_Type, ignore.case = TRUE))|> 
  mutate( subregion = toupper(County))

```

### Step 1

As gganimate extends the grammar of graphics as implemented by ggplot2 to include the description of animation. Therefore, we will need to generate a base map using ggplot2. Then, add a function of facet_wrap(Hospital_Category_RiskAdjustment) to track the average performance of each hospital category in each county across California.

### Step 2

Then, the gganimate library is used to animate the display of color by year (2019 – 2023). Here we define the transition time (in years), add a title and subtitle.

Note: 1. transition\_\*() defines how the data should be spread out and how it relates to itself across time. <br> 2. The comparison is depicted values and colors range from -1 (purple, worse than the national average) to 1 (yellow, better than the national average).

```{r}
#| title: Bloodstream infections (BSI) compare to national overtime
#| code-fold: true
#| warning: false
#| fig-width: 8
#| fig-height: 8


st <- mrsa_combine|>
    dplyr::select(subregion, Year, Comparison,Hospital_Category_RiskAdjustment)|>
  mutate(Year = as.integer(Year))|>
  group_by(subregion, Year,Hospital_Category_RiskAdjustment)|> 
  summarise(`national baseline` = round(mean(na.omit(as.numeric(Comparison))),2)) 

st <- left_join(st, unique(us_counties), by= "subregion", relationship = "many-to-many") 

a <-
st|> 

  ggplot(mapping = aes(x = long, y = lat, group = group, fill = `national baseline` ))+
              geom_polygon(color = "gray90", linewidth = 0.3) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_continuous(type = "viridis")+
  theme(axis.line=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid=element_blank())+
        scale_color_gradient(low = "green", high = "red")+ 
   facet_wrap(~Hospital_Category_RiskAdjustment)
            
            
a + transition_time(Year) +labs(title = "MRSA BSI incidence compares to the national baseline, Year: {frame_time}")

```

### Bonus

Also compatible with other ggplot graphs!

In this example we see trend plots of continuous variables infections cases and days of patient hospital stays between 2019 and 2023.

Did you notice the length of patient stay in a hospital is not necessarily positive correlated with numbers of reported cases (except for the acute care hospital category)?

```{r}
#| title: Patients Days vs Reported infection cases 
#| code-fold: true
#| warning: false
#| fig-width: 8
#| fig-height: 8


st <- left_join(mrsa_combine, unique(us_counties), by= "subregion", relationship = "many-to-many") 

p <-     
  st|> 
  ggplot(aes(x= Patient_Days, y= Infections_Reported))+
    geom_point(alpha = 0.5, show.legend = FALSE, color = "#167bb2") +
    scale_size(range = c(2, 12)) +
    theme_bw()+
    ggtitle(paste0("Length of Patients stay vs Reported infection cases")) +
    theme(#axis.text.x=element_blank(),
          axis.text=element_text(size=12),
          legend.position = "bottom",
          axis.line = element_line(colour = "black"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank())+
          facet_wrap(~Hospital_Category_RiskAdjustment)+
    xlab("Days")+ylab("Cases")+labs(caption="Produced by CF Lee")


p + transition_time(as.integer(Year))+  
  enter_fade() +
  exit_fade()+labs(title = "Year: {frame_time}")

```

## Resources

1.  gganimate [link](https://gganimate.com/)
2.  California Hospitals [data](https://data.chhs.ca.gov/dataset/methicillin-resistant-staphylococcus-aureus-mrsa-bloodstream-infections-bsi-in-california-hospitals).
