---
title: "Automated Reporting: A Game-Changer for Data Analysis "
author:
    name: Chris Lee
    url: https://clfee.github.io/
date: 2024-08-16
format: 
  html:
        html-math-method: katex
        grid: 
            margin-width: 350px
fig-width: 10
fig-height: 6
params:
  year: 2022
  record_date: 2024
editor_options: 
  chunk_output_type: console
execute: 
  echo: fenced
reference-location: margin
citation-location: margin
categories:
  - R 
  - Data organization
preview: "cover.PNG"
---

In the fast-paced world of data analysis, efficiency is key. Automated reporting is revolutionizing how we handle recurring data-rich reports, offering on-demand production and enhanced data exploration capabilities. This innovative approach not only saves time but also allows for deeper, more strategic insights. Parameterized reporting is a cutting-edge technique that enables simultaneous generation of multiple reports. This method not only boosts productivity but also significantly reduces errors associated with manual data handling.

**Goal :** The objective of this report is to demonstrate how to create automated custom reports with `Quarto` and `Purrr` in R to track patient days from public data.

## Real-World Application: California MRSA Data Analysis

To illustrate the power of automated reporting, let's consider a case study using data from the California [California Health and Human Services Open Data Portal](https://data.chhs.ca.gov/dataset/methicillin-resistant-staphylococcus-aureus-mrsa-bloodstream-infections-bsi-in-california-hospitals). The demo uses the data from All California general acute care hospitals are required to report Methicillin-resistant Staphylococcus aureus (MRSA) bloodstream infection (BSI) cases that occur following hospitalization. MRSA is a serious, contagious bacterial infection that starts on your skin. It’s a type of staph infection that resists most common antibiotics, making it especially dangerous. Without treatment, MRSA can be deadly.

## Step-by-Step Guide

### Step 1: Create a Single Report Template

We began by preparing a single report template using data from 2022. This template serves as the foundation for all subsequent reports. Let's prepare a single report as a template from the data in `r params$year`.

#### Healthcare Associated Infection Report, `r params$year`

```{r}
#| context: setup
#| include: false


# Load required packages
library(tidyverse)
library(ggplot2)

# Load data
url_2022 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/3ff809c6-c1e2-4107-be0c-f976521aadb2/download/mrsa_bsi_odp_2022.csv"
url_2021 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/23bef156-ae4d-4800-8512-d29660b2269f/download/cdph_mrsa_bsi_odp_2021.csv"
url_2019 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/62aead61-e320-4a74-b64c-1f65efa72b35/download/cdph_mrsa_bsi_odp_2019.csv"
url_2023 <-"https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/15c4962f-e357-4b03-9d24-655a2bc8030c/download/mrsa_bsi_odp_2023.csv"

data_2023 <- read_csv(url_2023)
data_2022 <- read_csv(url_2022)
data_2021 <- read_csv(url_2021)
data_2019 <- read_csv(url_2019)

# Combine data
mrsa_combine <- bind_rows(data_2023,data_2022, data_2021, data_2019)

# Remove redundant data
mrsa_combine <- unique(mrsa_combine)

# Data transformation
mrsa_combine <- mrsa_combine |>
  select(-State, -HAI, -Met_2020_Goal, -SIR_CI_95_Lower_Limit, -SIR_CI_95_Upper_Limit, -Comparison, -On_Track, -Notes) |>
  mutate(Months = as.numeric(ifelse(is.na(Months), 13, Months)),
         Year = as.numeric(ifelse(is.na(Year), 2021, Year)),
         Months = as.character(Months),
         #Year = as.character(Year),
         Infections_Reported = as.numeric(Infections_Reported))|>
  filter(!grepl("Pooled", Facility_Name, ignore.case = TRUE),
         !grepl("Pooled", Facility_Type, ignore.case = TRUE))
```

##### Results

**Figure 1:** Box plots of Standardized Infection Ratio (SIR) in `r params$year`.

```{r}
#| label: Boxplots
#| code-fold: true
#| warning: false

color_palette <- c("#167bb2","#2e3b4f","#7A6C5D","#2e4f42","#A54657","#624151","#16b29b","#002c47","#9d72d0"
)

 mrsa_combine|>
    filter(Year == params$year) |>
    filter(County %in% c("San Diego" ,"San Francisco", "Los Angeles", "Sacramento", "Orange","Yolo" ))|>
    ggplot(aes(x= Hospital_Type, y= SIR, fill=Hospital_Type))+
     geom_boxplot()+
    scale_fill_manual(values=color_palette)+
       theme(
      legend.position="none",
      plot.title = element_text(size=11),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    ) +
    ggtitle(paste0("Standardized Infection Ratio (SIR) in ",params$year)) +
    xlab("")+ylab("")+labs(caption="Produced by CF Lee")+
   facet_wrap(~County)
 
```

**Figure 2:** Patients Days vs Reported infection cases in `r params$year`

```{r}
#| code-fold: true
#| warning: false

 mrsa_combine$Months <- as.numeric(mrsa_combine$Months)

 mrsa_combine |> 
    filter(Year == params$year) |>
    mutate(Quarters = case_when(
                         Months > 0 & Months < 4 ~ 1,
                         Months > 3 & Months < 7 ~ 2,
                         Months > 6 & Months < 10 ~ 3 , 
                         Months > 9 & Months < 13 ~ 4 , 
                                .default = 4))|>
  # group_by(Quarters,Hospital_Category_RiskAdjustment, Year)|>
  #  summarise(mean = mean(na.omit(Infections_Reported)))|> 
    
    ggplot(aes(x= Patient_Days, y= Infections_Reported, 
               colour = as.integer(Quarters)))+
     geom_point(alpha = 0.7, show.legend = FALSE) +

    theme_bw()+
    ggtitle(paste0("Patients Days vs Reported infection cases in ",params$year)) +
    theme(#axis.text.x=element_blank(),
          axis.text=element_text(size=12),
          legend.position = "bottom",
          axis.line = element_line(colour = "black"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank())+
          facet_wrap(~Hospital_Category_RiskAdjustment)+
    xlab("Days")+ylab("Cases")+labs(caption="Produced by CF Lee")
```

Note: Patient Days is define as below. A count of the number of patients in the hospital during a time period, excluding IRFs and IPFs with their own CCN numbers. A rehabilitation unit (IRF) with its own CCN number reports separately from the rest of the hospital. To calculate patient days, the number of patients is recorded at the same time each day for each day of the month. At the end of the month, the daily counts are summed. See the NHSN website: http://www.cdc.gov/nhsn.

#### Step 2: Generate All Report Variations

After creating the single report, we used an R script to render the parameterized Quarto template with each defined parameter. This step allows for the simultaneous generation of all report variations, ensuring consistency and accuracy.To do this, write an R script to render the parameterized Quarto template with each of the defined parameters.

In our example the code looks like this,

```         
# 1. First create a dataframe 

data <- expand.grid(
  year = c(2019,2021:2023),
  record_date = Sys.Date(), 
  stringsAsFactors = FALSE)

df <- data |> 
  dplyr::mutate(
    output_format = "html",       # Output format (html, word, etc.)
    output_file = paste(          # Output file name
      year, "report.html",
      sep = "-"
    ),
    execute_params = purrr::map2( # Named list of parameters
      record_date, year, 
      \(record_date, year) list(record_date = record_date, year = year)
    )
  ) |> 
  dplyr::select(-c(record_date, year))

df

# 2. Use purrr::pwalk() to map over each row of the dataframe and render each report variation.

purrr::pwalk(
  .l = df,                      # Dataframe to map over
  .f = quarto::quarto_render,   # Quarto render function
  input = "demo_autoreport.qmd",       # Named arguments of .f
  .progress = TRUE              # Optionally, show a progress bar
)
```

By adopting this automated approach, analysts can shift their focus from report creation to data interpretation, leading to more impact decision-making and strategic planning.

Automatically generated reports, one after another!

{{< video demo.mp4 >}}

## Resources

1.  The raw data is obtained from California Health and Human Services Open Data Portal [datasets](https://data.chhs.ca.gov/dataset/methicillin-resistant-staphylococcus-aureus-mrsa-bloodstream-infections-bsi-in-california-hospitals)

2.  Full code [link](https://github.com/clfee)
