---
title: "Automated Custom reports - parameterized reporting "
author:
    name: Chris Lee
    url: https://clfee.github.io/
date: 2024-08-16
format: 
  html:
        html-math-method: katex
        grid: 
            margin-width: 350px
fig-width: 10
fig-height: 12
params:
  year: 2021
  record_date: 2024
editor_options: 
  chunk_output_type: console
execute: 
  echo: fenced
reference-location: margin
citation-location: margin
categories:
  - R 
  - Data organization
preview: "main.PNG"
---

Parameterized reporting is a technique that allows to generate multiple reports simultaneously. The technique also makes your work more accurate, as it avoids copy-and-paste errors. The objective of this report is to demonstrate how to create automated custom reports to track patient days from public data.

## Healthcare- Associated Infection Report, `r params$year`

Healthcare- Associated Infections can occur during healthcare. Methicillin-resistant Staphylococcus aureus (MRSA) is one of these diseases.MRSA is a serious, contagious bacterial infection that starts on your skin. It’s a type of staph infection that resists most common antibiotics, making it especially dangerous. Without treatment, MRSA can be deadly.

All California general acute care hospitals are required to report MRSA bloodstream infection (BSI) cases that occur following hospitalization so that CDPH can calculate the rates of MRSA BSI at each hospital and make that information available to the public (Health and Safety Code Section 1288.55).

## Demo- single report

Let's look at the data of Patient Days across the State in California in `r params$year`.

Note: Patient Days is define as below. A count of the number of patients in the hospital during a time period, excluding IRFs and IPFs with their own CCN numbers. A rehabilitation unit (IRF) with its own CCN number reports separately from the rest of the hospital. To calculate patient days, the number of patients is recorded at the same time each day for each day of the month. At the end of the month, the daily counts are summed. See the NHSN website: http://www.cdc.gov/nhsn.

## Demo- All variations at once

Once you are done with the single report. It is time to render all report variations. To do this, write an R script to render the parameterized Quarto template with each of the defined parameters.

In our example it looks like this,

```         
# 1. First create a dataframe 

data <- expand.grid(
  year = c(2019,2021:2023),
  record_date = Sys.Date(), 
  stringsAsFactors = FALSE)

df <- data |> 
  dplyr::mutate(
    output_format = "html",       # Output format (html, word, etc.)
    output_file = paste(          # Output file name
      year, "report.html",
      sep = "-"
    ),
    execute_params = purrr::map2( # Named list of parameters
      record_date, year, 
      \(record_date, year) list(record_date = record_date, year = year)
    )
  ) |> 
  dplyr::select(-c(record_date, year))

df

# 2. Use purrr::pwalk() to map over each row of the dataframe and render each report variation.

purrr::pwalk(
  .l = df,                      # Dataframe to map over
  .f = quarto::quarto_render,   # Quarto render function
  input = "demo_autoreport.qmd",       # Named arguments of .f
  .progress = TRUE              # Optionally, show a progress bar
)
```

## Data and code

The raw data is obtained from California Health and Human Services Open Data Portal [datasets](https://data.chhs.ca.gov/dataset/methicillin-resistant-staphylococcus-aureus-mrsa-bloodstream-infections-bsi-in-california-hospitals)

Full code [link](https://clfee.github.io/)

```{r}
#| context: setup
#| include: false

# Load required packages
#library(dplyr)
library(tidyverse)
#library(stringr)
library(ggplot2)

# Load data
url_2022 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/3ff809c6-c1e2-4107-be0c-f976521aadb2/download/mrsa_bsi_odp_2022.csv"
url_2021 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/23bef156-ae4d-4800-8512-d29660b2269f/download/cdph_mrsa_bsi_odp_2021.csv"
url_2019 <- "https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/62aead61-e320-4a74-b64c-1f65efa72b35/download/cdph_mrsa_bsi_odp_2019.csv"
url_2023 <-"https://data.chhs.ca.gov/dataset/a345bba7-25ed-4019-b6e3-0b597bea7368/resource/15c4962f-e357-4b03-9d24-655a2bc8030c/download/mrsa_bsi_odp_2023.csv"

data_2023 <- read_csv(url_2023)
data_2022 <- read_csv(url_2022)
data_2021 <- read_csv(url_2021)
data_2019 <- read_csv(url_2019)

# Combine data
mrsa_combine <- bind_rows(data_2023,data_2022, data_2021, data_2019)

# Remove redundant data
mrsa_combine <- unique(mrsa_combine)

# Data transformation
mrsa_combine <- mrsa_combine |>
  select(-State, -HAI, -Met_2020_Goal, -SIR_CI_95_Lower_Limit, -SIR_CI_95_Upper_Limit, -Comparison, -On_Track, -Notes) |>
  mutate(Months = as.numeric(ifelse(is.na(Months), 13, Months)),
         Year = as.numeric(ifelse(is.na(Year), 2021, Year)),
         Hospital_Cap = case_when(
           grepl("<125 Beds", Facility_Type) ~ 124,
           grepl(">250 Beds", Facility_Type) ~ 300,
           grepl("125-250", Facility_Type) ~ 185,
           grepl("Pediatric", Facility_Type) ~ 111,
           grepl("Major Teaching", Facility_Type) ~ 1449,
           grepl("Rehabilitation Unit", Facility_Type) ~ 301,
           grepl("Long-Term Acute Care", Facility_Type) ~ 301,
           grepl("Critical Access", Facility_Type) ~ 31,
           # Assigned random numbers to unknown
           TRUE ~ 99
         ),
         Patient_Stay = cut(      
           Patient_Days,
           breaks = quantile(Patient_Days, probs = seq(0, 1, by = 0.1), na.rm = TRUE),
           labels = LETTERS[1:10]),
         Months = as.character(Months),
         Year = as.character(Year),
         Infections_Reported = as.numeric(Infections_Reported))|>
  filter(!grepl("Pooled", Facility_Name, ignore.case = TRUE),
         !grepl("Pooled", Facility_Type, ignore.case = TRUE))
```

## Results

```{r}
#| label: Boxplots
#| code-fold: true
#| warning: false

color_palette <- c("#86C7ED","#167bb2","#2e3b4f","#7A6C5D","#2e4f42",
    "#A54657","#624151","#16b29b","#002c47"
)


 mrsa_combine|>
    filter(Year == params$year) |>
    ggplot(aes(x= Hospital_Type, y= Patient_Days, fill=Hospital_Type))+
     geom_boxplot()+
    scale_fill_manual(values=color_palette)+
       theme(
      legend.position="none",
      plot.title = element_text(size=11),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    ) +
    ggtitle(paste0("Patients Days in ",params$year)) +
    xlab("")+ylab("")+labs(caption="Produced by CF Lee")+
   facet_wrap(~County)
 
```
